Rhyme: A Query Language for Nested Data Structures

### 2024-12-13 paper.tex
finish section 2.1

To optimize performance, we apply advanced code motion and loop scheduling algorithms, enabling the generation of efficient program structures. These techniques ensure that the final implementation is both high-performing and adaptable to a variety of backend languages.\par
\iffalse
In this paper, we introduce the workflow of \rhyme language. We discuss how it supports both relational algebra and sparse/dense tensor computation in one framework. We also talk about the type system and type inference algorithm of \rhyme and how it enables efficient code generation for typed language backends as well as sparse tensor algebra.
\fi
In this paper, we present the workflow of the \rhyme language, highlighting its ability to unify relational algebra and sparse/dense tensor computations within a single framework. We detail the design of \rhyme's type system and its type inference algorithm, demonstrating how they facilitate efficient code generation for strongly typed language backends and support specialized computations for sparse tensor algebra.
\subsection{Paper Structure}
\iffalse
The paper is organized as follows: \par \par
In section \ref{workflow}, we will briefly go through the general workflow of the \rhyme compiler.\par
In section \ref{typesystem}, we present the type system of \rhyme as well as the type inference and checking algorithm.\par
In section \ref{sparse}, we will discuss how \rhyme supports specialized code generation for sparse tensor algebra, especially how it supports efficient co-iterating multiple sparse tensors.\par
In section \ref{evaluation}, we compare the generated code with TACO and explain the difference.\par
In section \ref{conclusion}, we summarize our work and draw some conclusions.
\fi
The structure of this paper is as follows:\par \par
Section \ref{workflow} provides an overview of the general workflow of the \rhyme compiler.\par
Section \ref{typesystem} introduces the type system of \rhyme, along with its type inference and checking algorithm.\par
Section \ref{sparse} delves into \rhyme's support for specialized code generation for sparse tensor algebra, with a focus on efficient co-iteration over multiple sparse tensors.\par
Section \ref{evaluation} evaluates the generated code by comparing it with TACO and highlights the differences.\par
Section \ref{conclusion} concludes the paper with a summary of our work and key takeaways.\par
\section{Workflow}\label{workflow}
\section{Type System}\label{typesystem}
\section{Sparse Tensor Algebra}\label{sparse}
\section{Evaluation}\label{evaluation}
\section{Conclusion}\label{conclusion}
%\newpage
\bibliographystyle{acm}
\bibliography{paper}
\end{document}
\endinput

>>>>

To optimize performance, we apply advanced code motion and loop scheduling algorithms, enabling the generation of efficient program structures. These techniques ensure that the final implementation is both high-performing and adaptable to a variety of backend languages.\par
\iffalse
In this paper, we introduce the workflow of \rhyme language. We discuss how it supports both relational algebra and sparse/dense tensor computation in one framework. We also talk about the type system and type inference algorithm of \rhyme and how it enables efficient code generation for typed language backends as well as sparse tensor algebra.
\fi
In this paper, we present the workflow of the \rhyme language, highlighting its ability to unify relational algebra and sparse/dense tensor computations within a single framework. We detail the design of \rhyme's type system and its type inference algorithm, demonstrating how they facilitate efficient code generation for strongly typed language backends and support specialized computations for sparse tensor algebra.
\subsection{Paper Structure}
\iffalse
The paper is organized as follows: \par \par
In section \ref{workflow}, we will briefly go through the general workflow of the \rhyme compiler.\par
In section \ref{typesystem}, we present the type system of \rhyme as well as the type inference and checking algorithm.\par
In section \ref{sparse}, we will discuss how \rhyme supports specialized code generation for sparse tensor algebra, especially how it supports efficient co-iterating multiple sparse tensors.\par
In section \ref{evaluation}, we compare the generated code with TACO and explain the difference.\par
In section \ref{conclusion}, we summarize our work and draw some conclusions.
\fi
The structure of this paper is as follows:\par \par
Section \ref{workflow} provides an overview of the general workflow of the \rhyme compiler.\par
Section \ref{typesystem} introduces the type system of \rhyme, along with its type inference and checking algorithm.\par
Section \ref{sparse} delves into \rhyme's support for specialized code generation for sparse tensor algebra, with a focus on efficient co-iteration over multiple sparse tensors.\par
Section \ref{evaluation} evaluates the generated code by comparing it with TACO and highlights the differences.\par
Section \ref{conclusion} concludes the paper with a summary of our work and key takeaways.\par
\section{Workflow}\label{workflow}
In this section, we will walk through the end-to-end workflow of \rhyme, demonstrating how the \rhyme query get compiled and generate efficient code to execute the query. Figure \ref{workflow} demonstrates the pipeline of \rhyme compiler.\par
\begin{figure}[H]
  \centering
      \includegraphics[scale=0.5]{figures/workflow.png}
      \caption{End-to-end workflow of \rhyme}
      \label{workflow}
  \end{figure}
\subsection{\rhyme Frontend}
\iffalse
\rhyme has multiple frontends like Pipe API, Textual API and Javascript API. \rhyme is a DSL embedded in Javascript, and as we mentioned before the \rhyme query can itself be a JSON object. Therefore we can use the JSON API to write query where the query is itself a JSON object. For example, if we want to group the sum of values of a dataset by its keys, we can just put \inline{data.*.key} as the key of the object query and put \inline{data.*.value} as the value as shown below. It is worth noting that since the \inline{sum(data.*.value)} is nested with \inline{data.*.key}, the logic variable in \inline{sum(data.*.value)} is bounded, the sum is actually a group-by sum. In addition to JSON API where the user can express the query in the same structure as the results, \rhyme also support Pipe API where the query is expressed as a sequence of transformations. The input wil pass through all the transformations to generate the final results. If we apply the logic variable \inline{.*} to a data source, the input will be iterated and a stream of its data will be passed for the following operation. The pipes are separated by \inline{|} symbol. The JSON API and Pipe API are equivalent and offers different ways to express the query for different application and scenarios. \par
\fi
\rhyme provides multiple frontends, including the Pipe API, Textual API, and JavaScript API. As a domain-specific language (DSL) embedded in JavaScript, \rhyme allows queries to be represented as JSON objects. This design enables users to construct queries directly in JSON format. For instance, to group the sum of values in a dataset by its keys, users can specify \inline{data..key} as the object query's key and \inline{data..value} as the value, as shown in the example below. Notably, because \inline{sum(data..value)} is nested within \inline{data..key}, the logic variable in \inline{sum(data.*.value)} is bound, making the operation a group-by sum.

In addition to the JSON API, where queries are expressed in the same hierarchical structure as their results, \rhyme also offers a Pipe API. The Pipe API allows queries to be defined as a sequence of transformations, where the input data passes through each transformation step to produce the final result. By applying the logic variable \inline{.*} to a data source, the input is iterated, and a stream of data is passed to subsequent operations. Transformations in the Pipe API are separated using the \inline{|} symbol.

The JSON API and Pipe API are equivalent in functionality but provide different methods for expressing queries, catering to various applications and scenarios.\par
\vspace*{1pt}
\noindent
\begin{minipage}{0.5\textwidth}\label{jsonapi}
\begin{lstlisting}
// JSON API: group the the sum of values by their keys
let query =  { data.*.key : sum(data.*.value) }
\end{lstlisting}
\end{minipage}%
\begin{minipage}{0.5\textwidth}
\begin{lstlisting}
//  Pipe API: group the sum of values by their keys
let key = rh`.data | .* | .key`
let query = rh`.data | .* | .value | group $\$${key}`
\end{lstlisting}
\end{minipage}
\vspace*{1pt}
\iffalse
The \rhyme query will be first parsed into an AST. The AST is a structured representation of the query and represented using JSON objects. \rhyme AST categorize operations into stateful operations (aggregation) and stateless operations (pure). The keys used for group-by will remain unchanged in the AST node. \rhyme will recursively translate the query into AST nodes and set the operator type and parameter correctly at each level.\par
In order to support generation of strongly typed languages like C/C++, \rhyme also allows user to provide optional type schemas of the input. As shown in Figure \ref{typeschema}, the user specify that the data is an object with u8 (8-bit unsigned integer) keys and string values. This type schema is crucial for our backend code generation as it provides type information for each sub-expressions of the query and allow us to generate more specialized code.
\fi
The \rhyme query is first parsed into an Abstract Syntax Tree (AST), a structured representation of the query expressed as JSON objects. The \rhyme AST categorizes operations into two types: stateful operations (e.g., aggregations) and stateless operations (e.g., pure functions). Keys used for group-by operations remain unchanged within the corresponding AST nodes. The \rhyme compiler recursively translates the query into AST nodes, assigning the appropriate operator types and parameters at each level.

To support code generation for strongly typed languages like C/C++, \rhyme allows users to optionally provide type schemas for the input data. As illustrated in Figure \ref{typeschema}, users can specify, for example, that the data consists of objects with 8-bit unsigned integer (u8) keys and string values. This type schema is essential for backend code generation, as it provides precise type information for each sub-expression of the query, enabling the generation of highly specialized and efficient code.
\begin{figure}[H]
  \centering
      \includegraphics[scale=0.4]{figures/typeschema.png}
      \caption{Example type schema of \rhyme}
      \label{typeschema}
  \end{figure}
\subsection{\rhyme Middle-end}
\subsection{\rhyme Backend}
\section{Type System}\label{typesystem}
\section{Sparse Tensor Algebra}\label{sparse}
\section{Evaluation}\label{evaluation}
\section{Conclusion}\label{conclusion}
%\newpage
\bibliographystyle{acm}
\bibliography{paper}
\end{document}
\endinput
